Characteristic polynomials of p-adic matrices

### 2017-02-04 charpoly/charpoly.tex
References

The characteristic polynomial is a fundamental invariant of a matrix: its roots
give the eigenvalues, and the trace and determinant can be extracted from its coefficients.
In fact, the best known division-free algorithm for computing determinants
over arbitrary rings \cite{kartofen:92a} does so using the characteristic polynomial.
Over $p$-adic fields, computing the characteristic polynomial is a key ingredient in
algorithms for counting points of varieties over finite fields. \todo{list some point counting references.}
\todo{Application to characteristic polynomials of elements of p-adic field extensions?}

@@

Let $M \in M_n(\Qp)$ be known at precision $O(p^m)$.
The simplest approach for computing the characteristic polynomial of $M$
is to compute $\det(X-M)$ either using recursive row expansion or various division free
algorithms \cites{\cite{seifullin:02a} \cite{kartofen:92a}}.  There are two issues
with these methods.  First, they are slower than alternatives that allow division,
requiring $O(n!)$, $O(n^4)$ and $\softO(n^{2 + \omega/2})$ operations.  Second,
while the lack of division implies that the result is accurate modulo $p^m$ as
long as $M \in M_n(\Zp)$, they still do not yield the optimal precision.

@@

\item in the model of floating-point arithmetic (see 
\cite{caruso-coursepadic}*{\S 2.3}), the number of correct digits of the 
$k$-th coefficient of the characteristic polynomial of $M$.
\end{enumerate}

>>>>

The characteristic polynomial is a fundamental invariant of a matrix: its roots
give the eigenvalues, and the trace and determinant can be extracted from its coefficients.
In fact, the best known division-free algorithm for computing determinants
over arbitrary rings \cite{kaltofen:92a} does so using the characteristic polynomial.
Over $p$-adic fields, computing the characteristic polynomial is a key ingredient in
algorithms for counting points of varieties over finite fields. \todo{list some point counting references.}
\todo{Application to characteristic polynomials of elements of p-adic field extensions?}

@@

Let $M \in M_n(\Qp)$ be known at precision $O(p^m)$.
The simplest approach for computing the characteristic polynomial of $M$
is to compute $\det(X-M)$ either using recursive row expansion or various division free
algorithms \cites{\cite{seifullin:02a} \cite{kaltofen:92a}}.  There are two issues
with these methods.  First, they are slower than alternatives that allow division,
requiring $O(n!)$, $O(n^4)$ and $\softO(n^{2 + \omega/2})$ operations.  Second,
while the lack of division implies that the result is accurate modulo $p^m$ as
long as $M \in M_n(\Zp)$, they still do not yield the optimal precision.

@@

\item in the model of floating-point arithmetic (see 
\cite{caruso:17a}*{\S 2.3}), the number of correct digits of the 
$k$-th coefficient of the characteristic polynomial of $M$.
\end{enumerate}
